---
title: "Introduction to optLump"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to optLump}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Categorical variables with many rare levels can often cause problems such as overfitting.
A potential solution is to lump these rare categories together, to
ensure every level meets a minimum sample size constraint.

Unlike naive approaches that simply lump the smallest levels into a
generic `Other` category, `optLump` uses information theory to
find the combinations that preserve the maximum amount of mutual
information from the original data distribution.
For an (abridged) explanation of mutual information, see `vignette("metrics")`.

The figure below shows the basic idea for an ordinal variable: levels whose
count falls below a threshold `q` (red dashed line) are absorbed into an
adjacent level.

```{r, eval=knitr::is_html_output(), echo=FALSE, message=FALSE, warning=FALSE}
library(plotly)
library(optLump)
library(dplyr)

counts <- c(12, 1, 8, 6, 1, 4, 2, 5)
orig_levels <- LETTERS[1:8]
x <- factor(rep(orig_levels, counts), levels = orig_levels, ordered = TRUE)

make_bars <- function(q) {
  lumped_x <- as.character(lump_ordinal(x, q))
  labels <- sapply(orig_levels, function(lv) lumped_x[which(as.character(x) == lv)[1]])
  data.frame(orig = orig_levels, pos = seq_along(orig_levels), lumped = labels, count = counts) |>
    group_by(lumped) |>
    summarise(x_pos = mean(pos), width = n() - 0.2, n = sum(count), .groups = "drop")
}

slider_step <- function(q) {
  agg <- make_bars(q)
  list(
    method = "update",
    args = list(
      list(x = list(I(agg$x_pos)), y = list(I(agg$n)), width = list(I(agg$width))),
      list(
        shapes = list(list(
          type = "line", x0 = 0, x1 = 1, xref = "paper",
          y0 = q, y1 = q,
          line = list(color = "red", dash = "dash", width = 2)
        ))
      )
    ),
    label = as.character(q),
    value = q
  )
}

init <- make_bars(6)

plot_ly() |>
  add_bars(x = I(init$x_pos), y = I(init$n), width = I(init$width),
           marker = list(color = "#7F77DD")) |>
  layout(
    xaxis = list(
      tickmode = "array",
      tickvals = I(seq_along(orig_levels)),
      ticktext = I(orig_levels),
      range = c(0.5, length(orig_levels) + 0.5)
    ),
    yaxis = list(title = "Count", range = c(0, 40)),
    bargap = 0,
    shapes = list(list(
      type = "line", x0 = 0, x1 = 1, xref = "paper",
      y0 = 6, y1 = 6,
      line = list(color = "red", dash = "dash", width = 2)
    )),
    sliders = list(list(
      active = 5,
      currentvalue = list(prefix = "Threshold q = "),
      steps = lapply(1:30, slider_step)
    ))
  ) |>
  config(displayModeBar = FALSE)
```

```{r, eval=!knitr::is_html_output(), echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(optLump)
library(dplyr)

counts <- c(12, 1, 8, 6, 1, 4, 2, 5)
orig_levels <- LETTERS[1:8]
x <- factor(rep(orig_levels, counts), levels = orig_levels, ordered = TRUE)
q <- 6

lumped_x <- as.character(lump_ordinal(x, q))
labels <- sapply(orig_levels, function(lv) lumped_x[which(as.character(x) == lv)[1]])
level_df <- data.frame(orig = orig_levels, pos = seq_along(orig_levels),
                       lumped = labels, count = counts)

before <- data.frame(xmin = seq_along(orig_levels) - 0.4,
                     xmax = seq_along(orig_levels) + 0.4,
                     n = counts, panel = "Before")
after  <- level_df |>
  group_by(lumped) |>
  summarise(xmin = min(pos) - 0.4, xmax = max(pos) + 0.4, n = sum(count), .groups = "drop") |>
  mutate(panel = sprintf("After (q = %d)", q))

rbind(before, after[, c("xmin", "xmax", "n", "panel")]) |>
  mutate(panel = factor(panel, levels = c("Before", sprintf("After (q = %d)", q)))) |>
  ggplot(aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = n)) +
  geom_rect(fill = "#7F77DD", colour = "white", linewidth = 0.3) +
  geom_hline(yintercept = q, colour = "red", linetype = "dashed") +
  scale_x_continuous(breaks = seq_along(orig_levels), labels = orig_levels,
                     limits = c(0.5, length(orig_levels) + 0.5)) +
  facet_wrap(~panel) +
  theme_minimal() +
  labs(y = "Count") +
  theme(panel.grid.major.x = element_blank(),
        strip.text = element_text(face = "bold"))
```

> **Note**
>
> This vignette only covers the primary usecase of optLump, looking purely at the distribution of
> the variable of interest. To take into account more information, see `vignette("supervised")`.

# Handling Different Types of Variables
```{r, include = FALSE}
# Generate simulated dataset
set.seed(0)
n <- 100

country_of_origin <- factor(sample(
  c("US", "Canada", "Mexico", "France", "Germany", "China", "Japan", "South Korea", "Vietnam"),
  n, replace = TRUE
))

degrees <- c("<High School", "High School", "Bachelor's", "Master's", "PhD")
education_level <- ordered(sample(degrees, n, replace = TRUE, prob = c(0.1, 0.4, 0.3, 0.15, 0.05)), levels = degrees)

diagnosis <- factor(sample(
  c("None", "Diabetes", "Hypertension", "Heart Disease"),
  n, replace = TRUE, prob = c(0.4, 0.25, 0.25, 0.1))
)

data <- data.frame(
    outcome = rpois(n, 1),
    education_level = education_level,
    diagnosis = diagnosis,
    country_of_origin = country_of_origin
    )
```

Suppose we are given some data containing a response variable `outcome`, based
on some predictors, which are categorical:
```{r}
summary(data)
```
This means that we would be trying to model `outcome` as a function of
the predictors.
The goal we want to achieve is to lump the categories of the predictors together
in order to reduce the dimensionality of the problem.
The treatment for each type of variable is different, and optLump provides different methods for each type of variable.

```{r setup}
library(optLump)
```

## Ordinal Variables

The simplest case is that of ordinal variables, which are variables that have a natural ordering of the categories.
In our example, the variable `education_level` is an ordinal variable.
We might decide that the smaller levels, `<High School` and `PhD` are too
small to be useful, and
that we want to lump them together with other levels.
We would do this by specifying a threshold value, in this instance 15,^[For help choosing
this threshold, see `threshold_diagnostic()`.]
which is the minimum number of observations that we want in each level after lumping.
```{r}
lumped_levels <- lump_ordinal(data$education_level, 15)
summary(lumped_levels)
```

We see that the algorithm has lumped the level `<High School` with `High School`,
and the level `PhD` with `Master's`.
Of course, the interpretation of the levels has changed (the new levels should
perhaps be interpreted as "High school or less" and "Postgraduate"),
but we see that all levels meet the threshold of 15 observations now.

## Nominal Variables

For nominal variables, we lose the ordering and thus we can't rely on the computer to automatically
know what levels can be lumped together.
The simplest, and most common approach is to simply allow all
levels to be lumped with all other levels.
This is the default behaviour:
```{r}
lumped_levels <- lump_nominal(data$diagnosis, 15)
summary(lumped_levels)
```

A less common situation is when we have some prior knowledge about
which levels can be lumped together.
Suppose we know based on domain knowledge that `Heart Disease` should
not be lumped with `Hypertension`, since they are too different.
This information can also be encoded in the preference graph,^[The preference
graph is a graph that encodes what lumpings are allowed; it has an edge between
two levels if and only if those levels are allowed to be lumped together.]
by disallowing the corresponding pairing.
Notice that the preference graph is taken in as its adjacency matrix.
```{r}
preference_graph <- adjacency_from_edge_list(
  levels(data$diagnosis),
  disallow = list(c("Heart Disease", "Hypertension"))
)
lumped_levels <- lump_nominal(data$diagnosis, 15, preference_graph)
summary(lumped_levels)
```
Of course, if the preference graph is more sparse, corresponding
to a situation in which only a small number of pairings are admissible,
it would make more sense to build the graph the other way around, by
starting with an empty graph and adding edges for the pairings that are allowed.

### Hierarchical Variables

An important subclass of nominal variables are hierarchical variables,
where we have some structure inherited a hierarchy.
A common example -- and the one demonstrated here -- is that of geographical
variables.
We know, a priori, that countries belong to continents, naturally
clustering countries together, so it makes more sense to lump countries
together that exist within the same continent.
It is simple to pass this knowledge in:
```{r}
clusters <- list(
    c("US", "Canada", "Mexico"),
    c("France", "Germany"),
    c("China", "Japan", "South Korea", "Vietnam")
)
lumped_levels <- lump_hierarchical(data$country_of_origin, 15, clusters)
summary(lumped_levels)
```

The reason this function exists separately from `lump_nominal()` -- which
can achieve the exact same behaviour by passing an appropriate preference graph --
is, in addition to the ergonomic gains, that it allows certain speedups that are
not possible in the general case.

# Dplyr Integration

```{r, message = FALSE}
library(dplyr)
```

The package works very naturally with dplyr pipelines.
For instance, we can easily apply the `lump_ordinal()` function to the `education_level` variable in our dataset:
```{r}
data |>
  mutate(education_level = lump_ordinal(education_level, 15)) |>
  slice_head(n = 10)
```

# Computation and Large Datasets

Due to inherent computational constraints,^[being that the nominal
problem is NP-hard] `lump_nominal()` might be very slow.
In fact, the algorithm has $O\left(2^{2^m}\right)$ runtime, where
$m$ is the number of levels.
To partially remedy this, `lump_hierarchical()` is substantially faster
when it can be used, having $O\left(\frac{m}{c}2^{2^c}\right)$, where
$c$ is the number of levels in the largest cluster.
In our example, we would have $c = 4$, since we have 4 Asian countries.
Since $c \leq m$, this is a substantial improvement.

When even this is too slow, we can use `lump_nominal_heuristic()`, which
approximates a solution in polynomial time, but does not guarantee an
optimal solution.

Note that `lump_ordinal()` does not suffer from the same issue and runs
in polynomial time.
