---
title: "Supervised Lumping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Supervised Lumping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In the introductory vignette (`vignette("optLump")`) we lumped a categorical
variable by looking only at its own distribution.
This is the right thing to do when the variable is all we have.
More often, however, we are lumping a predictor because we eventually want to model
some outcome based on it.
In that case we do not really care about the predictor on its own; we care about
what it tells us about the outcome.

Supervised lumping takes this into account.
Instead of preserving the mutual information between the variable before and after
lumping, it preserves the mutual information between the lumped variable and an
outcome that we supply.^[For an explanation of mutual information, see `vignette("metrics")`.]
In practice this means that two small levels are lumped together when they say
similar things about the outcome, rather than simply when doing so keeps the
distribution balanced.

Each of the supervised functions corresponds to one of the unsupervised functions
from `vignette("optLump")`.
They take the same arguments, with one addition: the `outcome` that supervises the
lumping.

```{r setup}
library(optLump)
```

# Lumping with an Outcome

```{r, include = FALSE}
set.seed(1)
n <- 200

n_per <- c(Minimal = 22, Mild = 60, Moderate = 14, Severe = 48, Critical = 56)
severity <- ordered(rep(names(n_per), n_per), levels = names(n_per))

admission <- factor(
  sample(c("GP", "Specialist", "Emergency", "Self-referral", "Walk-in clinic"),
         n, replace = TRUE, prob = c(0.30, 0.25, 0.25, 0.10, 0.10)),
  levels = c("GP", "Specialist", "Emergency", "Self-referral", "Walk-in clinic")
)

hospital <- factor(
  sample(c("St Thomas'", "Guy's", "Whittington", "Royal Free", "Homerton"),
         n, replace = TRUE, prob = c(0.28, 0.24, 0.20, 0.18, 0.10)),
  levels = c("St Thomas'", "Guy's", "Whittington", "Royal Free", "Homerton")
)

severity_effect  <- c(Minimal = -1.6, Mild = -1.4, Moderate = -1.5, Severe = 1.0, Critical = 1.2)
admission_effect <- c(GP = 0, Specialist = 0.2, Emergency = -0.6,
                      "Self-referral" = 1.0, "Walk-in clinic" = -1.0)
logit <- severity_effect[as.character(severity)] + admission_effect[as.character(admission)]
readmitted <- factor(rbinom(n, 1, plogis(logit)))

los_mean <- c(Minimal = 2.5, Mild = 3, Moderate = 2.5, Severe = 8, Critical = 10)[as.character(severity)]
length_of_stay <- round(rgamma(n, shape = 6, scale = los_mean / 6), 1)

patients <- data.frame(severity, admission, hospital, readmitted, length_of_stay)
```

Suppose we are given some data on hospital admissions.
For each patient we know a triage `severity`, the `admission` route they came in
through, and the `hospital` that treated them.
What we would like to model is whether the patient was `readmitted` within a month,
as well as their `length_of_stay` in days.

```{r}
summary(patients)
```

As before, we want to lump the categories of the predictors together so that every
level meets some minimum sample size.
The difference is that we now want to do this without throwing away what the
predictors tell us about the outcome.

## Ordinal Variables

Just like in the unsupervised case, `severity` is ordinal, so we are only allowed
to lump levels that are next to each other in the ordering.
Suppose we want every level to contain at least 30 patients.
Lumping without the outcome looks only at the counts:

```{r}
summary(lump_ordinal(severity, 30))
```

To meet the threshold, the algorithm has lumped the small `Moderate` level in with
`Severe`.
This is a sensible choice if all we know are the counts, but it is an unfortunate
one here: `Moderate` patients are rarely readmitted, whereas `Severe` patients
usually are.

```{r}
round(tapply(readmitted == "1", severity, mean), 2)
```

If we supervise the lumping with the outcome, the algorithm can take this into
account.
We simply pass the outcome in as the second argument:

```{r}
summary(lump_ordinal_supervised(severity, readmitted, 30))
```

This time `Moderate` is lumped with the lower-severity levels that it actually
resembles, and `Severe` is left on its own.
The resulting levels make a lot more sense if we are about to use them to predict
readmission.

### Continuous Outcomes

The outcome does not have to be discrete.
If we give it a numeric outcome such as `length_of_stay`, the same function instead
preserves the mutual information between the lumped variable and that continuous
outcome, estimated using a nearest-neighbour approach [@ross2014].
By default, optLump infers the type of the outcome from its class---a factor is
treated as discrete, a numeric vector as continuous---so there is nothing extra to
specify:

```{r}
summary(lump_ordinal_supervised(severity, length_of_stay, 30))
```

Since longer stays go hand in hand with higher severity, we end up with the same
grouping as before.
Of course, if the automatic choice is ever wrong -- say the outcome is stored as
integers but is really categorical -- we can set it by hand through the
`outcome_mode` argument.

## Nominal Variables

For nominal variables we lose the ordering, so in principle any two levels can be
lumped together.
`admission` is such a variable.
Asking once more for at least 25 patients per level, the unsupervised method lumps
together the two smallest levels:

```{r}
summary(lump_nominal(admission, 25))
```

`Self-referral` and `Walk-in clinic` are indeed the two rarest routes.
The trouble is that they disagree about the outcome: self-referred patients are
readmitted often, whereas walk-in patients hardly ever are.

```{r}
round(tapply(readmitted == "1", admission, mean), 2)
```

The supervised version keeps them apart, and instead lumps each rare level in with
a larger one that behaves like it:

```{r}
summary(lump_nominal_supervised(admission, readmitted, 25))
```

Just as in the unsupervised case, we can supply a preference graph through the
`adj_matrix` argument to forbid certain pairings, and a continuous outcome is
handled exactly as it was for ordinal variables.
See `vignette("optLump")` for how to build a preference graph.

### Hierarchical Variables

When the levels come with a known grouping, `lump_hierarchical_supervised()` can
take advantage of it, just like `lump_hierarchical()` does in the unsupervised
case.
Suppose our five hospitals belong to two networks, and that we only want to lump
patients together within the same network:

```{r}
networks <- list(
  c("St Thomas'", "Guy's"),
  c("Whittington", "Royal Free", "Homerton")
)
summary(lump_hierarchical_supervised(hospital, readmitted, 25, clusters = networks))
```

Within each network, the rarest hospitals are lumped together in the way that holds
on to the most information about readmission.
Unlike the ordinal and nominal versions, the hierarchical function only supports
discrete outcomes.

# Computation and Large Datasets

The supervised functions behave computationally much like their unsupervised
counterparts.
`lump_ordinal_supervised()` runs in time cubic in the number of levels.
`lump_nominal_supervised()`, on the other hand, solves an NP-hard problem, so it is
only practical for a modest number of levels.
When the data has a hierarchy, `lump_hierarchical_supervised()` is substantially
faster, since it only ever works on one cluster at a time.
For more on these trade-offs, and on the heuristics available when the nominal
problem becomes too slow, see `vignette("optLump")`.

# References
